\documentclass[12pt, letter, openany]{book}
\usepackage{format}

\newcommand{\jissue}{1}
\newcommand{\jyear}{2025}


\begin{document}

\frontmatter

\vspace*{\fill}
\begin{center}
Copyright © 2025 by the authors.

All articles published in Academus are open access under the Creative Commons Attribution-NonCommercial 4.0 International License (CC BY-NC 4.0).

Authors retain copyright and grant the journal a non-exclusive license to publish and distribute their work.
\end{center}

\vspace*{\fill}

\chapter{Editor's Note}

Welcome to the inaugural issue of Academus, a student-run journal for the humanities and social sciences. We started Academus because we want to engage today’s youth with big questions— questions about knowledge and power, justice and community, technology and tradition. We noticed a gap: these brilliant minds were without a clear place to publish, debate, and build on one another’s ideas. This issue is our first attempt to fill that gap. We want to create a home for serious, accessible scholarship by high-school authors who aren’t waiting to be asked to join the conversation when they get to college. 

What follows is intentionally broad. The papers range across philosophy, political theory, history, economics, and public policy, and they share two traits we prize: argumentative clarity and concern for real-world stakes. We’re grateful to the authors and peer reviewers who trusted a brand-new journal with their work. We hope the pieces spark more inquiry, more disagreement, and most importantly, more writing. 

We open with Jacob Gong’s “A Binary Theory of Collapse,” which models civilizational failure as the interaction of anthropological and environmental forces, then presses the unsettling claim that contemporary climate dynamics shift that model from diagnostic to predictive. The essay’s synthesis, from Tainter to modern climate literature, invites readers to test theory against history and present policy. 

Joseph Huang’s “Evaluating Governmental Control on Social Media Speech” enters today’s most fraught public square. Weighing feasibility, abuse risks, and economic effects, the paper argues that state regulation is a poor instrument against online falsehoods compared with third-party fact-checking and scalable media literacy. In essence, he asks: what works without eroding liberal commitments? 

In “A Critique of the College Board Company,” Reed Chan traces the growth of a nonprofit into a de facto gatekeeper for admissions and curriculum, pairing historical analogy with contemporary antitrust concerns. Whether you agree with every charge, the piece raises a timely question about public goods administered by private actors. 

Justin Lee’s “A Fertility Crisis” surveys demographic decline across cultural, economic, and biological drivers, then rejects narrow pronatalism in favor of strengthening the underlying economic conditions that make family formation feasible. It’s an essay that refuses easy answers and keeps the policy horizon wide. 

Michael Gkatzimas’s “A History of Racism and a Case for Affirmative Action” distinguishes interpersonal from systemic racism and evaluates affirmative action under utilitarian and Kantian lenses. The argument lands on a modest but firm conclusion: imperfect tools can still be the fairest response to unfair histories. 

Kevin Nguyen’s “China’s Lost Children, Its Dying Economy, and the Economic Implications” reads the long shadow of the One-Child Policy across labor markets, savings behavior, housing, and global spillovers. The piece exemplifies our aim to connect institutional choices to worldwide consequences. 

In “On Humility,” Yichen Wang stages the Feldman–Kelly peer-disagreement debate and defends doxastic humility: when equally competent thinkers conflict, suspension of judgment is often the rational (and cooperative) move. It’s meta-epistemology with everyday implications for learning together. 

We close with Alexander Lian’s “Balancing Free Speech and Fair Elections: A Historical Analysis of U.S. Campaign Finance,” which moves from Tillman to Buckley to Citizens United and beyond, mapping how rules about money, speech, and parties keep rearranging the democratic game. The history doubles as a civics prompt: what kind of guardrails best protect both voice and vote?

Finally, I want to mention that we’re indebted to the teachers, mentors, and reviewers who made this first issue possible, and to the authors who trusted us with their work. If something here makes you want to argue, write it down, and send it in. May these pages invite careful thinking, charitable disagreement, and scholarship that carries beyond these covers.

\medskip

\emph{Vertias $\cdot$ Ratio $\cdot$ Sapientia.}

\medskip

\noindent Sincerly,

Sam Cao, Editor-in-Chief

\tableofcontents

\mainmatter

\begin{refsection}[refs/binary]
\input{source/binary}
\nocite{*}
\printbibliography[title={References}, heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/evaluating]
\input{source/evaluating}
\nocite{*}
\printbibliography[title=References,
heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/critique]
\input{source/critique}
\nocite{*}
\printbibliography[title=References, heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/fertility]
\input{source/fertility}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/racism]
\input{source/racism}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/china]
\input{source/china}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/humility]
\input{source/humility}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/campaign]
\input{source/campaign}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\end{document}
