\documentclass[12pt, a4paper, openany]{book}
\usepackage{format}

\newcommand{\jissue}{2}
\newcommand{\jyear}{2025}

\begin{document}

\frontmatter

\vspace*{\fill}
\begin{center}
Copyright © 2025 by the authors.

All articles published in Academus are open access under the Creative Commons Attribution-NonCommercial 4.0 International License (CC BY-NC 4.0).

Authors retain copyright and grant the journal a non-exclusive license to publish and distribute their work.
\end{center}

\vspace*{\fill}

\chapter{Editor's Note}

Welcome to the second issue of Academus. This issue circles a common concern: how institutions and ideas shape the lives we actually live. From election rules to classroom pressures, from classic debates in moral philosophy to the early American presidency, our authors test arguments not only for elegance, but for impact. We’re grateful to the writers and reviewers who made this volume possible, and we hope these pages invite you to argue back: with care, with courage, and in writing.

We open with Alexander Lian’s “Can Ranked Choice Voting Deliver Meaningful Electoral Reform in the U.S.?” A set of case studies (Alaska statewide races, San Francisco 2018, Oakland 2010) shows RCV reliably blunting the spoiler effect and nudging campaigns toward broader coalitions—even as it reconfigures, rather than erases, strategic behavior.

Charles Schäfer’s “The American Dream’s Flawed Promise of Education: The Issue of Academic Stress” traces how rising academic competition burdens adolescents—psychologically, physically, and socially—and how supportive school cultures, mental-health resources, and resilience practices can convert pressure into growth. The argument insists that access and ambition mean little without belonging and well-being.

In “Against Moral Intuitionism,” Celine Wang challenges our confidence in “self-evident” moral seemings. Drawing on reliability worries, deep disagreement, and Mackie’s queerness objection, the paper presses readers to ask not just \emph{what} we feel is right, but \emph{why} those feelings should count as knowledge.

Sarah Kim’s “The Political Economy of Farm Subsidies” argues that U.S. agricultural supports are regressive at home and distortive abroad—socializing costs, concentrating benefits, and weakening developing-country advantages—while proposing a freer, fairer alternative that takes both consumers and farmers seriously.

Sam Cao’s “The Asymmetry of Moral Obligations” uses Michael Moehler’s multilevel social-contract framework to separate duties to contemporaries (grounded partly in shared norms) from duties to future people (grounded in prudence and bargaining). The piece contrasts Kyoto’s prescriptivism with Paris’s pragmatic architecture to show how cooperation can scale under moral pluralism.

Jeff Cao’s “Rejection of Cognitivism” (after T.\ Nguyen) reframes trust not as belief but as an \emph{unquestioning attitude} that rationally delegates deliberation under cognitive limits. By disentangling trust from belief, the paper explains how evidentialism can survive those everyday moments when trust and surface evidence collide.

With “Washington’s Neutrality Proclamation and the Birth of Executive Authority,” John Wang revisits 1793. Through language, law, and politics, the essay shows how Washington’s carefully ambiguous “friendly and impartial” stance both expanded the practical reach of the presidency and forged a public-facing model of executive leadership.

We close with Suresh Raina’s “A Study of Ticket Scalping,” which defends resale markets as allocatively efficient and—counter to intuition—often equitable. By walking through simple models and real frictions, the piece invites readers to separate outrage at bots from the economics of getting scarce seats to those who value them most.

Finally, I want to mention that we’re indebted to the teachers, mentors, and reviewers who made this second issue possible, and to the authors who trusted us with their work. If something here makes you want to argue, write it down, and send it in. May these pages invite careful thinking, charitable disagreement, and scholarship that carries beyond these covers.

\medskip
\emph{Veritas · Ratio · Sapientia.}

\medskip

\noindent Sincerely,

 Sam Cao, Editor-in-Chief

\tableofcontents

\mainmatter

\begin{refsection}[refs/rcv]
\input{source/rcv}
\nocite{*}
\printbibliography[title={References}, heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/american]
\input{source/american}
\nocite{*}
\printbibliography[title=References,
heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/moralint]
\input{source/moral}
\nocite{*}
\printbibliography[title=References, heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/farm]
\input{source/farm}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/asymmetry]
\input{source/asymmetry}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/rejection]
\input{source/rejection}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/washington]
\input{source/washington}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\begin{refsection}[refs/tickets]
\input{source/tickets}
\nocite{*}
\printbibliography[title=References,heading=subbibliography]
\end{refsection}

\end{document}
